Goal Document: Translating Typing Rules into JastAdd

### 2023-02-09 GoalDocument.tex
goaldoc: Minor adjustments as per feedback

\section{Background}
In programming languages there are many types of data, and for any real world use case there will be constraints upon these types, such as what types of input a function can contain.
Type checking is the method of verifying, either at runtime (dynamic checking) or compile time (static checking), that these constraints are held.

JastAdd is a meta-compilation system, allowing one to work with and extend abstract syntax trees with great flexibility and is extensively used within the SDE group at LTH.
However, implementing a type checker in JastAdd by hand currently requires a lot of boilerplate code, slowing down iteration.
Typing rules offer a condensed, standardised syntax for defining simple languages, allowing fast iterations with minimal work, but these rules then typically need to be implemented by hand.

@@

The goal of this project is to translate the concisely formatted typing rules into a working type-checker in JastAdd.
We'll start with a basic implementation able to handle simple syntax-directed rules.
Once this is working, there are two possible directions to take the project.
One is to support object-oriented languages, by adding support for subtyping and ad hoc polymorphism.
The other is to support a more functional approach by implementing a Hindley-Milner style type system.
Which of these routes to take will be decided based on the experience of implementing the basic system.

@@

\section{Previous work}
\begin{itemize}
  \item
    André Pacak, Sebastian Erdweg, and Tamás Szabó. 2020. A systematic approach to deriving incremental type checkers. Proc. ACM Program. Lang. 4, OOPSLA, Article 127 (November 2020), 28 pages. https://doi.org/10.1145/3428195 \\
    \url{https://szabta89.github.io/publications/inca-typechecking.pdf} \\
    A paper about implementing a compiler, capable of deriving incremental type checking from typing rules.
    Similar goal to ours, but with a significantly different compilation target, meaning we will likely differ in architecture and what trade-offs are made.

@@

\section{Evaluation}
The first step of evaluating the project is to see what language features it supports.
Once the program is capable of compiling a basic language, we can also compare it to existing handwritten (or written ourselves) JastAdd specifications for those same languages.
Here we could evaluate the size of the finished specification as well as the performance of the type checking.

An additional point of comparison is the Datalog paper, to which we could compare supported features as well as output size and performance.
%  Case studies, implementation possible by hand or already finished implementations
%  Compare with handwritten implementations:
%    - Spec size
%    - Performance
%  Compare with datalog paper

>>>>

\section{Background}
In programming languages there are many types of data, and for any real world use case there will be constraints upon these types, such as what types of input a function can contain.
Type checking is a method of verifying, either at runtime (dynamic checking) or compile time (static checking), that these constraints are held.

JastAdd is a meta-compilation system, allowing one to work with and extend abstract syntax trees with great flexibility. % and is extensively used within the SDE group at LTH.
However, implementing a type checker in JastAdd by hand currently requires a lot of boilerplate code, slowing down iteration.
Typing rules offer a condensed, standardised syntax for defining simple languages, allowing fast iterations with minimal work, but these rules then typically need to be implemented by hand.

@@

The goal of this project is to translate the concisely formatted typing rules into a working type-checker in JastAdd.
We will start with a basic implementation able to handle simple syntax-directed rules.
Once this is working, there are two possible directions to take the project.
One is to support object-oriented languages, by adding support for subtyping and ad hoc polymorphism.
The other is to support a more functional approach by implementing a Hindley-Milner style type system.
Which of these routes to take will be decided based on the experience of implementing the basic system.

@@

\section{Related work}
\begin{itemize}
  \item
    André Pacak, Sebastian Erdweg, and Tamás Szabó. 2020. A systematic approach to deriving incremental type checkers. Proc. ACM Program. Lang. 4, OOPSLA, Article 127 (November 2020), 28 pages. https://doi.org/10.1145/3428195 \\
    \url{https://szabta89.github.io/publications/inca-typechecking.pdf} \\
    A paper about implementing a compiler, capable of deriving incremental type checking from typing rules.
    Similar goal to ours, but with a significantly different compilation target, meaning we will likely differ in architecture and what trade-offs are made.

@@

\section{Evaluation}
The first step of evaluating the project is to see what language features it supports. % What are the challenges? What properties of typesystems make them difficult/easy to embed? Efficiency, spec size
Once the program is capable of compiling a basic language, we can also compare it to existing handwritten (or written ourselves) JastAdd specifications for those same languages.
Here we will evaluate the size of the finished specification as well as the performance of the type checking.

An additional point of comparison is Pacak et al.'s \emph{A systematic approach to deriving incremental type checkers}, to which we will compare supported features as well as output size and performance.
%  Case studies, implementation possible by hand or already finished implementations
%  Compare with handwritten implementations:
%    - Spec size
%    - Performance
%  Compare with datalog paper
